---
title: "Change in Number of Deaths From AIDS"
output:
  html_document:
    df_print: paged
editor_options:
  chunk_output_type: console
---


```{r echo=FALSE, message=FALSE, warning=FALSE}
library(dplyr)
library(tidyr)
library(ggplot2)
library(reshape2)


options(scipen=999)
```


```{r echo=FALSE, message=FALSE, warning=FALSE}
# Read data into a dataframe
master <- read.csv('Indicator_Annual number of AIDS deaths.csv', 
                   check.names = F)
```

#The Basics: Univariate Analyses


```{r echo=FALSE, message=FALSE, warning=FALSE}
#Create a Tidy dataset to work from
tidymaster <- gather(master, "year", "n", '1990':'2011', na.rm=TRUE)

```



```{r echo=FALSE, message=FALSE, warning=FALSE}
summary(tidymaster)
summary(master)
```

Summaries of tidied and untidied data. 

The majority of countries have low death numbers relative to each other; 
the median is always very far below the mean. For this project, I'm generally
more interested in countries with numbers above the mean.


```{r echo=FALSE, message=FALSE, warning=FALSE}
# creates a dataframe of all countries with at least one year above the mean
above_mean <- subset(master, master$`1990` > 2150 | master$`1991` > 2687 |
  master$`1992` > 3300 | master$`1993` > 4064 | master$`1994` > 4876 | 
  master$`1995` > 5786 | master$`1996` > 6712 | master$`1997` > 7490 | 
  master$`1998` > 8423 | master$`1999` > 9462 | master$`2000` > 10383 |
  master$`2001` > 11335 | master$`2002` > 12280 | master$`2003` > 12985 |
  master$`2004` > 13438 | master$`2005` > 13598 | master$`2006` > 13502 |
  master$`2007` > 13040 | master$`2008` > 12443 | master$`2009` > 11709 |
  master$`2010` > 10434 | master$`2011` > 9706)

# tidies the above-mean dataframe
tidy.above_mean <- gather(above_mean, "year", "n", '1990':'2011', na.rm=TRUE)

``` 


```{r echo=FALSE, message=FALSE, warning=FALSE}
summary(above_mean)
```

This is a summary the dataset when it's made up only of countries that have at 
least one year where they were above the mean. In other words, excluding 
countries that are consistently below the mean.

# Examining More Than One Variable



```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height = 10, fig.width = 10}
ggplot(aes(x=n), data=tidymaster) +
  geom_histogram() +
  facet_wrap(~ Country) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

```

Faceting the data by country shows that there is at least some kind of change 
over time in many countries, even those that stay well below the median.


```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x=n), data=tidymaster) +
  geom_histogram(bins=50) +
  facet_wrap(~ year)
```

These graphs are the count of yearly death totals broken out by year for the 
entire dataset, not just those countries above the mean.

Notable elements include: starting in about 1999, more countries have higher 
death rates.

2010 and 2011 have significantly fewer records than in previous years.  

Although the introduction of AZT in the 90's began the dramatic decrease of
 rate of deaths from AIDS, the highest single-year rates are found in the
  late 2000's. This could be due to better reporting/communication, or an 
  indication of global inequalities in access to medicine.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x=n), data=tidy.above_mean) +
  geom_histogram(bins=50) +
  facet_wrap(~ year)

```

The same graph as above, but using the dataset of countries with at least one
year above the mean.

Although most countries are still reporting deaths on the lower end of the scale,
 a little more variation is visible, particularly after the late 90's, as 
 compared to the dataset of all countries.


```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x=n), data=tidy.above_mean) +
  geom_histogram(bins=100)

ggplot(aes(x=n), data=tidy.above_mean) +
  geom_histogram(bins=100) +
  facet_wrap(~year)

ggplot(aes(x=n), data=tidy.above_mean) +
  geom_histogram(bins=100) +
  facet_wrap(~Country)
```

Histograms for number of cases in all countries that have at least one year
above the mean for that year.


```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x=year, y=n), data=tidy.above_mean) +
  geom_point() +
  facet_wrap(~Country)
```

Despite the introduction of AZT in the early 90's, many countries saw increases
in reported deaths throughout the 90's. 

# Trends over two decades

```{r}
write.csv(above_mean, 'above_mean.csv')
```


```{r echo=FALSE, message=FALSE, warning=FALSE}
master2 <- subset(master, select = c('Country', '1990', '2000'))
master2 <- na.omit(master2)

summary(master2)

master2<-subset(master2, master2$`1990` > 2150)

ggplot(data=master2, aes(x=Country, y=master2$`1990`)) + 
  geom_bar(stat='identity')
```

This graph shows the number of deaths from AIDS in 1990, for
countries with deaths above the mean in 1990.


```{r echo=FALSE, message=FALSE, warning=FALSE}

master2long <- melt(master2, id.var = "Country")

ggplot(master2long, aes(x = Country, y = value, fill = variable)) + 
  geom_bar(stat = "identity", position = "dodge") 

```

Comparing 1990 to 2000, again with only the set of countries above the mean in
1990


```{r echo=FALSE, message=FALSE, warning=FALSE}

master3 <- subset(master, select = c('Country', '1990', '2000', '2010'))
master3 <- na.omit(master3)

master3<-subset(master3, master3$`1990` > 2150)

master3long <- melt(master3, id.var = "Country")

ggplot(master3long, aes(x = Country, y = value, fill = variable)) + 
  geom_bar(stat = "identity", position = "dodge") 

```

Tracing number of deaths from AIDS from 1990, 2000, and 2010, for countries 
above the mean number of deaths in 1990. 

Out of 12 countries surveyed, 10 saw an increase from 1990 to 2000, while only 3
 saw an increase between 2000 and 2010.

